---
title: "ENVS193DS Final"
author: Owen Eckweiler
date: 6/11/2025
editor: source
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

# Repository Link

<https://github.com/eckdogg/ENVS-193DS_spring-2025_final.git>


# Set up

```{r}
library("tidyverse")
library("here")
library("janitor")
library("tibble")
library("DHARMa")
library("MuMIn")
library("ggeffects")
library("scales") # reading in packages

sst <- read_csv(here("data", "SST_update2023.csv")) # creating object with csv for problem 2
nest_boxes <- read_csv(here("data", "occdist.csv")) # creating object with csv for problem 3
nest_boxes_clean <- janitor::clean_names(nest_boxes) # cleaning the nest_boxes object
```


# Problem 1. Research writing

## a. 

In part 1, they used a Pearson's r correlation test. In part 2, they used a one-way ANOVA test.

## b. 

Two additional tests that could be included are a Tukey's HSD test and an eta-squared test. Tukey's HSD will tell us what groups are significantly different from each other. In this example there are 5 groups (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands). An eta-squared test will give us the effect size between groups. Effect size will tell us how large the difference is between groups. 

## c.

There is a difference in average nitrogen load (kg year<sup>-1</sup>) between sources (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands) (p = 0.02). We found a large difference ($\eta^2$ = effect size) between sources in average nitrogen load (one-way ANOVA results). On average urban land tended to have larger nitrogen loads than atmospheric deposition (Tukey's HSD results), fertilizer (Tukey's HSD results), wastewater treatment (Tukey's HSD results), and grasslands (Tukey's HSD results).


# Problem 2. Data visualization

## a.

```{r}
sst_clean <- sst |> # cleaning data from sst data frame
  mutate(date = ymd(date), # creating date objects
         year = factor(year(date), levels = 2018:2023), # turning year variable into a factor
         month = month(date, label = TRUE, abbr = TRUE)) |> # creating a month column
  filter(!is.na(year)) |> # filtering out rows without a year value
  group_by(year, month) |> # grouping by year and month
  summarize(mean_monthly_sst = round(mean(temp, na.rm = TRUE), 1),
            .groups = "drop") # taking the mean sst for each month

sst_clean |> 
  slice_sample(n = 5) # displaying 5 random rows of data

str(sst_clean) # displaying the data structure
```

## b.

```{r}
ggplot(sst_clean, aes(x = month, y = mean_monthly_sst, group = year, color = year)) + # using sst_clean data frame and setting values
  geom_line(size = 0.8) + # line geometry
  geom_point(size = 1.5) + # point geometry
  labs( # axis labels
    x = "Month", # x axis label
    y = "Mean monthly sea surface temperature (°C)", # y axis label
    color = "Year" # coloring based on year
  ) +
  scale_color_manual( # manually setting color scale
    values = c(
      "2018" = "#deebf7", # color for 2018 (lightest shade)
      "2019" = "#c6dbef", # color for 2019
      "2020" = "#9ecae1", # color for 2020
      "2021" = "#6baed6", # color for 2021
      "2022" = "#3182bd", # color for 2022
      "2023" = "#08519c"  # color for 2023 (darkest shade)
    )
  ) +
  scale_y_continuous(limits = c(12, 20)) + # only displaying values between 12 and 20 on y axis
  theme_minimal(base_size = 14) + # setting base font size to 14 points
  theme(
    legend.position = c(0.00, 1.00),  # legend position
    legend.justification = c("left", "top"),  # pin legend to top-left
    plot.margin = margin(t = 10, r = 10, b = 10, l = 30), # plot margin
    axis.title.y = element_text(margin = margin(r = 10)), # margin for y axis
    panel.grid.major = element_blank(),  # removing major gridlines
    panel.grid.minor = element_blank(),  # removing minor gridlines
    panel.background = element_blank(),  # removing background
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)  # setting border color and size
  )
```


# Problem 3. Data analysis

## a.

A 1 represents the presence of a species and a 0 represents the absence of a species. For example a 1 underneath the sp column represents the presence of a swift parrot in a nest box.

## b.

In this study, the nest boxes were built for swift parrot habitat restoration. However in the location where the nest boxes were built there are two other species that compete with swift parrots for nest boxes (tree martins and common starlings).

## c.

The two years compared in the study are 2016 and 2019, which were subsequent breeding seasons. Nest boxes were originally established in 2016 and were used again in 2019. 

## d.

| Model number | Season | Distance to Forest Edge | Model Description | 
|:------------:|:---:|:---------------:|:-------:|-----------------------------|  
| 0            |        |                          | No predictors (null model) |
| 1            |  X     |       X                 | All predictors (saturated model) | 
| 2            |  X     |                         | Season |   
| 3            |       |       X                  | Distance to Forest Edge |  

## e.

```{r}
# model 0: null model
model0 <- glm(sp ~ 1, # formula
  data = nest_boxes_clean, # data frame
  family = "binomial" # family
)

# model 1: saturated model
model1 <- glm(sp ~ edge_distance + season, # formula
  data = nest_boxes_clean, # data frame
  family = "binomial" # family
)

# model 2: season
model2 <- glm(sp ~ season, # formula
  data = nest_boxes_clean, # data frame
  family = "binomial" # family
)

# model 3: edge distance
model3 <- glm(sp ~ edge_distance, # formula
  data = nest_boxes_clean, # data frame
  family = "binomial" # family
)
```

## f.

```{r}
plot(
  simulateResiduals(model0)
) # Displaying diagnostic plots for model0

plot(
  simulateResiduals(model1)
) # Displaying diagnostic plots for model1

plot(
  simulateResiduals(model2)
) # Displaying diagnostic plots for model2

plot(
  simulateResiduals(model3)
) # Displaying diagnostic plots for model3
```

## g.

```{r}
AICc(model0,
     model1, # best model: All predictors (saturated model). The predictors are distance from forest edge and season. The response variable is swift parrot occupancy.
     model2,
     model3) |>
  # arranging output in descending order of AIC
  arrange(AICc)
```

## h.

```{r}
nest_boxes_clean$season <- as.factor(nest_boxes_clean$season) # setting season variable as a factor

mod_preds <- ggpredict(model1, terms = c("edge_distance [all]", "season")) # using ggpredict with edge distance and season as predictors

ggplot() +
  geom_point(data = nest_boxes_clean, # data frame
             aes(x = edge_distance, # x axis variable
                 y = sp, # y axis variable
                 color = season), # coloring based on season
             size = 3, # size of points
             alpha = 0.4) + # transparency of points
  geom_ribbon(data = mod_preds, # data frame
              aes(x = x, # x axis variable
                  ymin = conf.low, # lower limit of confidence interval
                  ymax = conf.high, # upper limit of CI
                  fill = group), # coloring based on group
              alpha = 0.4) + # transparency
  geom_line(data = mod_preds, # data frame
            aes(x = x, # x axis variable
                y = predicted, # y axis variable
                color = group), # coloring based on group
            linewidth = 1) +  # width of line
  scale_y_continuous("Probability of box occupancy", # y axis label
                   limits = c(0, 1), # y axis limits
                   breaks = seq(0, 1, by = 0.2), # increments
                   labels = label_percent(accuracy = 1)) + # showing percentages on y axis
  scale_color_manual(name = "Season", # line colors
                   values = c("2016" = "darkblue", "2019" = "lightgreen")) +
  scale_fill_manual(name = "Season", # ribbon colors
                   values = c("2016" = "darkblue", "2019" = "lightgreen")) +
  labs(x = "Distance from the forest edge (m)", # x axis label
       title = "Swift Parrot") + # title
  theme_minimal(base_size = 14) + # minimal theme
theme(
  panel.grid = element_blank()) # removing gridlines

```

## i.

**Figure 1. Swift parrots are more likely to be found near forest edges.** Data from (Stojanovic, Dejan et al. (2021). https://doi.org/10.5061/dryad.83bk3j9sb).
Points represent observations of swift parrot presence at individual nest boxes at various distances from forest edges (m). Points at 100% on the y axis represent swift parrot presence and points at 0% represent a lack of presence. Lines represent the predicted probability of swift parrot presence based on distance from forest edge. Ribbons represent the upper and lower limits of confidence intervals (CI = 95%). Colors represent seasons (blue: 2016, green: 2019).

## j.

```{r}
nest_boxes_clean$season <- as.factor(nest_boxes_clean$season)

ggpredict(model1, terms = c("edge_distance [0]", "season"))

ggpredict(model1, terms = c("edge_distance [900]", "season"))
```

## k.

The predicted probability of a swift parrot occupying a nest box at 0 meters from the forest edge was much higher than at 900 meters for both seasons. However the probability of swift parrot occupation at both distances was higher in 2016 than in 2019. Distance from forest edge and probability of swift parrot occupancy have a negative relationship (probability decreases as distance increases). This relationship can be explained by the fact that tree martins occupied the most nest boxes in the study and were most likely to use boxes far from forest edges. As a result swift parrots favored nest boxes near forest edges due to decreased competition.


# Problem 4. Affective and exploratory visualizations

## a.

- My exploratory visualizations were created using code while my affective visualization was created using digital pencil. My exploratory visualizations were composed of points on a graph while my affective visualization is an artistic representation of box plots as buildings.
- All of my visualizations have axes and axis labels. They also both have some sort of geometry on the plot itself.
- My exploratory visualizations examined steps based on day of the week and maximum temperature while my affective visualization examined floors climbed on school days and non school days. There were not any obvious patterns in the exploratory visualizations because they were only based on four different data points. In my affective visualization, floors climbed were higher on non school days than on school days.
- For my affective visualization, An recommended that I use windows on the buildings to represent something about my data. As a result, I drew windows with different colors that corresponded to a specific number of cumulative floors climbed.

## b.

I completed this component in workshop.